Estimating the incidence of SARS-CoV-2 from prevalence data

### 2023-09-28 incidence-prevalence.tex
tweaking

\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
It was first developed in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome)~\autocite{brookmeyerMethod}.
HIV is a virus transmitted sexually or through blood, and causes the disease AIDS.
AIDS has killed around 40 million people; at its peak, around 2 million people died from AIDS each year~\autocite{unaids2023}.
In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses and the distribution of the time between infection and AIDS diagnosis~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
\todo{explain how AIDS diagnosis is "prevalence" data (is it?)}

Three processes are involved in backcalculation: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
The infection process describes how individuals become infected.
The prevalence process describes the relationship between incidence and prevalence.
The observation process describes the relationship between prevalence and the observed data.

>>>>

\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
It was first developed in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome)~\autocite{brookmeyerMethod}.
HIV is transmitted sexually or through blood.
It causes the disease AIDS.
AIDS has killed around 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the distribution of the time between infection and AIDS diagnosis~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
\todo{explain how AIDS diagnosis is "prevalence" data (is it?)}

Three processes are involved in backcalculation: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
The infection process determines incidence.
The prevalence process describes the relationship between incidence and prevalence.
The observation process describes the relationship between prevalence and the observed data.

The infection process is a point process, where each infection is an event.
The simplest plausible model is a time-inhomogenous Poisson process.
That is, a Poisson process where the intensity varies over time.
The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic.

Two features of a Poisson process can be violated for epidemics.
Epidemics commonly exhibit overdispersion and have feedback loops.
Overdispersion means that the variance in the number of infections in a time interval is higher than its mean.
In a Poisson process, counts in a time interval have a Poisson distribution.
Therefore, the mean and variance are equal.
Feedback loops mean that the stochastically low or high numbers of infections, possibly due to overdispersion, affect the future number of infections.
In a Poisson process, the intensity (mean number of events per unit time) is not influenced by the realisation of the process.
Both of these features violate the assumptions of a Poisson process.

A common cause of overdispersion is super-spreading events.
\todo{ref}
A super-spreading event is an event where many more people than expected are infected.
Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion.
\todo{insert a SARS-CoV-2 superspreading example}
Overdispersion could be incorporated by modelling the counts in a given time interval as a negative binomial, rather than Poisson, distribution.
The mean of these distributions would be equal, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
These distributions are related in that the negative binomial tends to Poisson as $k\to\infty$.
Generally, $k<1$ indicates significant levels of overdispersion.
Estimates for SARS-CoV-2 indicate $k \approx 0.1$, and can be increased by interventions such as lockdowns.
However, the negative binomial distribution loses much of the mathematical convenience of a Poisson distribution.
For a negative binomial process, the count's standard deviation scales on the order of the mean.
Conversely, for a Poisson process, the standard deviation is on the order of the square root of the mean.
A particular implication of this is that the coefficient of variation (the standard deviation divideed by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.

The independence assumption is violated because if the number of infections is stochastically high, this feeds back to the process intensity at future times.
This can be seen simply in the event of a super-spreading event.
If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
These additional infections will propagate the epidemic further, increasing the intensity at future times.
